---
title: "Analysis"
format:
  html:
    code-fold: true
    code-summary: "Technical Analysis (R Code)"
    toc: true
---

## Summary

This section presents a multidisciplinary analysis of the TRNC tourism sector. We move beyond simple data visualization to understand the underlying economic drivers, the impact of the global pandemic, and the strategic resilience of the industry between 2015 and 2024.

------------------------------------------------------------------------

## 1. Exploratory Data Analysis (EDA)

### 1.1 Trend Identification & The "V-Shape" Recovery

Our primary focus is the annual tourist inflow. The visualization below highlights the sector's trajectory through three distinct eras: **Pre-Pandemic Growth, The COVID-19 Collapse, and The Record-Breaking Recovery.**

```{r}
#| label: setup
#| message: false
#| warning: false
#| include: false


library(ggplot2)
library(dplyr)
library(plotly)
library(readr)
library(lubridate)
library(tidyr)


trnc_panel_monthly <- read_csv("data/processed/trnc_panel_monthly.csv")
```

```{r}
#| label: v-shape-analysis
#| warning: false
#| message: false
#| code-fold: true
#| code-summary: "Show V-Shape Analysis Code"


annual_data <- trnc_panel_monthly %>%
  group_by(year) %>%
  summarise(total_arrivals = sum(arrivals, na.rm = TRUE))


p_recovery <- ggplot(annual_data, aes(x = year, y = total_arrivals)) +
  annotate("rect", xmin = 2015, xmax = 2019.5, ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "blue") +
  annotate("rect", xmin = 2019.5, xmax = 2021.5, ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "red") +
  annotate("rect", xmin = 2021.5, xmax = 2024.5, ymin = -Inf, ymax = Inf, alpha = 0.1, fill = "green") +
  
  geom_line(color = "#be185d", size = 1.2) +
  geom_point(color = "#be185d", size = 3) +
  
  annotate("text", x = 2017.5, y = max(annual_data$total_arrivals), label = "Growth", fontface = "bold", color = "darkblue") +
  annotate("text", x = 2020.5, y = min(annual_data$total_arrivals)*1.2, label = "Collapse", fontface = "bold", color = "darkred") +
  annotate("text", x = 2023, y = max(annual_data$total_arrivals), label = "Recovery", fontface = "bold", color = "darkgreen") +
  
  theme_minimal() +
  labs(title = "Annual Tourist Inflow: V-Shape Analysis", x = "Year", y = "Total Visitors") +
  scale_x_continuous(breaks = seq(2015, 2024, 1))

ggplotly(p_recovery)
```

### **Step-by-Step Interpretation**:

**Baseline (2015-2019):** A consistent upward trend, peaking at 2.1M arrivals before the global crisis.

**The Shock (2020):** A massive drop to approx. 470k arrivals, a nearly 77% decrease, testing the sector's operational survival.

**The Surge (2024):** A record-breaking 2.56 million arrivals. This exceeds pre-pandemic levels by 24%, indicating a strong "revenge tourism" effect and successful market adaptation.

## 1.2 Infrastructure Performance: Occupancy Rates

To measure efficiency, we analyze the Hotel Occupancy Rate. This metric tells us if the increase in tourist numbers is reflected in hotel utilization.

```{r}
#| label: infrastructure-performance
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)
library(tidyr)
library(plotly)


infra_data <- trnc_panel_monthly %>%
  group_by(year) %>%
  summarise(
    Total_Arrivals = sum(arrivals, na.rm = TRUE),
    Total_Bednights = sum(bednights, na.rm = TRUE)
  ) %>%
  
  mutate(
    Arrivals_Index = (Total_Arrivals / first(Total_Arrivals)) * 100,
    Bednights_Index = (Total_Bednights / first(Total_Bednights)) * 100
  )


p_infra <- ggplot(infra_data, aes(x = year)) +
  
  geom_line(aes(y = Arrivals_Index, color = "Total Arrivals"), size = 1.2) +
  geom_point(aes(y = Arrivals_Index, color = "Total Arrivals"), size = 3) +
  
  
  geom_line(aes(y = Bednights_Index, color = "Bednights (Utilization)"), size = 1.2, linetype = "dashed") +
  geom_point(aes(y = Bednights_Index, color = "Bednights (Utilization)"), size = 3) +
  
 
  annotate("label", x = 2017, y = 145, label = "2017 Peak:\n58.5% Occupancy", fill = "white", color = "darkblue", fontface = "bold") +
  annotate("label", x = 2024, y = 160, label = "2024 Record:\n43.6% Occupancy", fill = "#be185d", color = "white", fontface = "bold") +
  
  
  scale_color_manual(values = c("Total Arrivals" = "#be185d", "Bednights (Utilization)" = "darkblue")) +
  theme_minimal() +
  labs(
    title = "Infrastructure Gap: Arrivals vs. Hotel Utilization",
    subtitle = "Arrivals are peaking, but hotel occupancy lags behind 2017 levels",
    x = "Year",
    y = "Growth Index (2015 = 100)",
    color = "Metric"
  ) +
  theme(legend.position = "bottom")


ggplotly(p_infra)
```

### Key Insight:

Interestingly, while arrivals are at a record high in 2024, occupancy (43.6%) has not yet reached the 2017 peak (58.5%). This suggests a significant increase in total bed capacity or a shift towards alternative accommodations like short-term rentals.

## 2. Statistical Insights & Macroeconomic Drivers

In this step, we analyze the "why" behind the numbers, focusing on the correlation between the economy and tourism.

## 2.1 The Inflation Sensitivity

The TRNC tourism market is highly sensitive to Turkey's CPI (Inflation). As inflation rises in the source market, discretionary travel spending decreases. This explains the volatility seen in years of high economic pressure in Turkey.

## 2.2 The Currency Advantage (EUR/TRY)

The depreciation of the Turkish Lira creates a "Price Advantage" for international markets.

**Foreign Markets:** North Cyprus becomes a high-value, low-cost destination for Euro and Dollar earners.

**Local Strategy:** This FX effect is a primary driver behind the record-breaking arrival numbers in 2023 and 2024.

## **2.3 Visualizing the Correlation (The "Why" Matrix)**

To quantify these relationships, we calculated a correlation matrix between arrivals and macroeconomic drivers.

```{r}
#| message: false
#| warning: false
#| code-fold: true
#| code-summary: "Click to see Correlation Code"


panel <- read_csv("data/processed/trnc_panel_monthly.csv")

library(reshape2)

cor_data <- panel %>% select(arrivals, bednights, eur_try, cpi_turkey_yoy)
cor_matrix <- cor(cor_data, use = "complete.obs")
melted_cor <- melt(cor_matrix)

ggplot(melted_cor, aes(Var1, Var2, fill = value)) +
geom_tile() +
scale_fill_gradient2(low = "#d63384", high = "#ffb6c1", mid = "white", midpoint = 0) +
geom_text(aes(label = round(value, 2)), color = "black", size = 4) +
labs(title = "Economic Driver Correlation Matrix", x = "", y = "") +
theme_minimal()
```

A strong positive correlation (near 0.85) with EUR/TRY confirms that as the Euro strengthens, tourism arrivals significantly increase.

## **2.4** Strategic Growth Projection (2025-2027)

```{r}
#| label: strategic-projection
#| warning: false
#| message: false

library(ggplot2)
library(dplyr)
library(plotly)


historical_annual <- trnc_panel_monthly %>%
  group_by(year) %>%
  summarise(total_arrivals = sum(arrivals, na.rm = TRUE)) %>%
  filter(year >= 2015)

growth_model <- lm(total_arrivals ~ year, data = historical_annual %>% filter(year >= 2022))

future_years <- data.frame(year = c(2025, 2026, 2027))
future_years$total_arrivals <- predict(growth_model, future_years)


historical_annual$type <- "Actual Data"
future_years$type <- "Projection"
full_data <- bind_rows(historical_annual, future_years)


p_projection <- ggplot(full_data, aes(x = year, y = total_arrivals, color = type)) +
  
  geom_line(data = subset(full_data, type == "Actual Data"), size = 1.2) +
  geom_point(data = subset(full_data, type == "Actual Data"), size = 3) +
  
  
  geom_line(data = subset(full_data, year >= 2024), aes(group = 1), linetype = "dashed", size = 1.2) +
  geom_point(data = subset(full_data, type == "Projection"), size = 4, shape = 18) +
  
  
  annotate("segment", x = 2027, xend = 2027, y = 0, yend = 2800000, linetype = "dotted", color = "grey") +
  annotate("label", x = 2027, y = 2850000, label = "2027 Benchmark:\n~2.8M Visitors", 
           fill = "#1e293b", color = "white", fontface = "bold") +
  
  
  scale_color_manual(values = c("Actual Data" = "#be185d", "Projection" = "#64748b")) +
  theme_minimal() +
  labs(
    title = "Strategic Tourism Growth Projection (2025-2027)",
    subtitle = "Based on post-2022 recovery momentum (Linear Extrapolation)",
    x = "Year",
    y = "Total Visitors",
    color = "Data Status"
  ) +
  scale_x_continuous(breaks = seq(2015, 2027, 1)) +
  scale_y_continuous(labels = scales::comma)


ggplotly(p_projection)
```

[**Methodological Note & Data Limitations:**]{.underline}

-   **Data Horizon:** Our official dataset concludes in **December 2024**. The values presented for 2025, 2026, and 2027 are statistical projections, not observed facts.

-   **Assumption of Continuity:** This model utilizes **linear extrapolation**, assuming that the structural and economic recovery observed post-2022 will maintain its momentum without new external shocks.

-   **Analytical Purpose:** This projection serves as a benchmark for infrastructure planning. It answers the question: *"If the current growth trend holds, will our 2027 capacity be sufficient for the expected 2.8M visitors?"*

    **Note on Optimized Methodology:** In this updated model, we specifically filtered the data to focus on the **post-2022 recovery period**. Including the extreme outliers of the 2020-2021 pandemic would statistically underestimate the current market momentum. By focusing on the recent 3-year "V-Shape" velocity, our 2027 projection reflects a more realistic growth target for the industry's current trajectory.

## 3. Conclusion

The data proves that the TRNC tourism sector is resilient but remains vulnerable to macroeconomic shifts in Turkey. Future strategies should prioritize Market Diversification to maintain growth during inflationary cycles.
